Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-19 article/article.Rmd
Commit 2024-09-19 12:14
Commit 2024-09-19 19:08

The cloud-free reference values discussed above are based on the climatological AOD
and WC; hence they cannot describe accurately the long-term variation of
$\text{GHI}_\text{ref}$ due to long-term changes in the two atmospheric constituents,
mainly AOD.  As reported by @Natsis2023, there is a long-term brightening effect in
the GHI data of Thessaloniki for the period 1993 -- 2023, which for cloud-free data
was attributed to long-term changes in aerosol effects.  Therefore, an adjustment of
the $\text{GHI}_\text{ref}$ during the period of study was made using RTM simulations
to account for the long-term variations of the AOD.  High quality AOD data with a
Cimel sun-photometer at Thessaloniki start only in 2003, while spectral AOD
measurements using direct-sun observations with a MKIII Brewer spectrophotometer are
available for the period 1997 -- 2017. The Brewer AOD data are taken sporadically
during each day and are less dense compared to the AERONET data. By comparing monthly
AOD data at $340\,\text{nm}$ of the two instruments for the common periods of
operation we adjusted the Brewer data and filled the missing months of the Cimel
record with Brewer data.  Using the extended monthly time series of AOD at
$340\,\text{nm}$, as well as monthly climatological values of the Ångström exponent
and constant WC of $15.6\,\text{mm}$ derived from the Cimel record, we simulated with
the RTM the cloud-free GHI at SZA of $55^\circ$ for each month in the period 1997 --
2023. The SZA of $55^\circ$ was chosen as representative of all days in the year to
get a rough estimate of the annually averaged change in cloud-free GHI. A
second-degree polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols:
\begin{equation}
\Delta(\text{GHI}) [\%] = -12170 + 12.05 \cdot x + -0.002981 \cdot x^2 \label{eq:AODchange}
\end{equation}
where $y$ is the date as a decimal fraction of the year.  The resulting change in
$\text{GHI}_\text{ref}$ ranges between $-1.88$ in 1994 and $-0.19\,\%$ in the end of
2023, peaking at $0.23\,\%$ in the mid-2020. Finally, we applied these long-term
changes of Equation\nobreakspace{}\ref{eq:AODchange} to the $\text{GHI}_\text{ref}$,
to create a more realistic representation of the cloud-free irradiance for the entire
period of study. For the period 1994 -- 1996 where no AOD data are available, we
assumed that the changes in GHI follow the same polynomial fit.

@@

```{r group-2d, echo=FALSE, fig.cap="Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1994 -- 2023. The logarithmic color scale denotes the frequency of the respective groups. Note that 19 pints with duration higher than 60 min are not shown."}
knitr::include_graphics("../images/P-groups-bin2d-1.pdf")
```

>>>>

The cloud-free reference values discussed above are based on the climatological AOD
and WC; hence they cannot describe accurately the long-term variation of
$\text{GHI}_\text{ref}$ due to long-term changes in the two atmospheric constituents,
mainly AOD.  As reported by @Natsis2023, there is a long-term brightening effect in
the GHI data of Thessaloniki for the period 1993 -- 2023, which for cloud-free data
was attributed to long-term changes in aerosol effects.  Therefore, an adjustment of
the $\text{GHI}_\text{ref}$ during the period of study was made using RTM simulations
to account for the long-term variations of the AOD.  High quality AOD data with a
Cimel sun-photometer at Thessaloniki start only in 2003, while spectral AOD
measurements using direct-sun observations with a MKIII Brewer spectrophotometer are
available for the period 1997 -- 2017. The Brewer AOD data are taken sporadically
during each day and are less dense compared to the AERONET data. By comparing monthly
AOD data at $340\,\text{nm}$ of the two instruments for the common periods of
operation we adjusted the Brewer data and filled the missing months of the Cimel
record with Brewer data.  Using the extended monthly time series of AOD at
$340\,\text{nm}$, as well as monthly climatological values of the Ångström exponent
and constant WC of $15.6\,\text{mm}$ derived from the Cimel record, we simulated with
the RTM the cloud-free GHI at SZA of $55^\circ$ for each month in the period 1997 --
2023. The SZA of $55^\circ$ was chosen as representative of all days in the year to
get a rough estimate of the annually averaged change in cloud-free GHI. A
second-degree polynomial fit was applied to the simulated yearly averaged GHI to
derive the long-term change in GHI due to aerosols:
\begin{equation}
\Delta(\text{GHI}) [\%] = -12170 + 12.05 \cdot x + -0.002981 \cdot x^2 \label{eq:AODchange}
\end{equation}
where $x$ is the date as a decimal fraction of the year.  The resulting change in
$\text{GHI}_\text{ref}$ ranges between $-1.88$ in 1994 and $-0.19\,\%$ in the end of
2023, peaking at $0.23\,\%$ in the mid-2020. Finally, we applied these long-term
changes of Equation\nobreakspace{}\ref{eq:AODchange} to the $\text{GHI}_\text{ref}$,
to create a more realistic representation of the cloud-free irradiance for the entire
period of study. For the period 1994 -- 1996 where no AOD data are available, we
assumed that the changes in GHI follow the same polynomial fit.

@@

```{r group-2d, echo=FALSE, fig.cap="Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1994 -- 2023. The logarithmic color scale denotes the frequency of the respective groups. Note that 19 points of groups with duration higher than 60 min are not shown."}
knitr::include_graphics("../images/P-groups-bin2d-1.pdf")
```
